Numerical simulation of bubble deformation \\ and breakup under simple linear shear flows

### 2024-11-27 Revised-paper/Paper/Bubble-deformation-Rev.tex
microgravity papers

\begin{abstract}
\textcolor{red}
{
For the first time, either experimentally or numerically, numerical simulations are presented for the deformation and breakup of a bubble in liquids undergoing simple linear shear flow in the low Capillary number, high Reynolds number regime.  Numerical results are obtained using a projection method for incompressible two-phase flow. The method represents interfaces using the sharp interface coupled level set and Volume-Of-Fluid (CLSVOF) method.   To verify our numerical algorithm and provide a basis for comparison, we also present simulation results that compare with previous, more popular, drop deformation and rupture simulations.  The numerical results reveal that the shear-induced bubble deformation and breakup process differs significantly from that of the analogous drop system under similar flow conditions.  The first distinguishing factor for bubble breakup is that the magnitude of shear flow necessary for causing rupture is much larger than that in the drop system case.  In other words, a larger Reynolds number is needed to induce bubble breakup.  The second distinct feature of the bubble system versus the drop system is that the bubble does not maintain a stable deformed shape as the system parameters approach the critical ``breakup'' Reynolds number.  It is asserted that the differences in morphology for a bubble undergoing breakup versus a drop in the same process can be attributed to the density and viscosity ratio of the corresponding two-phase flow systems.  The critical conditions for bubble breakup with respect to the Reynolds and capillary numbers are determined for several cases. }
\end{abstract}

@@

Bubble dynamics in shear flow, including breakup, is critically important for various \textcolor{blue} {gas-liquid} scientific and engineering processes.  We refer the reader to the following experimental studies relating to bubble deformation in foaming processes, microfluidic devices, microbubbles in the blood circulation system, materials processing, pharmaceuticals, minerals engineering, bubble column reactor, underwater projectiles, polyethylene devolatilization, microfluidics, food aeration, and cosmetics\cite{ChuFinBouAtaHamPug19,MulTobDreFisWin08,BenRodFauPinFerPerGarMirLim18,DreSai15,EFTEKHARI2021837,WANG2023108105,doi:10.1021/acs.langmuir.1c01814,yoshikawa2010bubble,CanedoETAL,GAO2022103212,WONG2012417,PolyethyleneDevo,FRENSE2024120579,lohse2018bubble,schluter2021small,SANOGO2023112478}.  In particular, it is the study of bubble deformation as it pertains to high-performance plastics applications that motivate this work.

@@

In previous experimental or computational studies on the motion of bubble deformation in a simple shear flow \cite{CanedoETAL,RusMan02, MulTobDreFisWin08}, only findings for bubble deformation under very low $Re$ number conditions ($Re \ll 1$) have been reported.  This is understandable since a low Reynolds number matrix fluid mitigates the effect of gravity on distorting the comparison with the drop case.  In the results reported in \cite{CanedoETAL}, the $x-z$ cross section of the deforming bubble was close to spherical in their experiments.  Also in the experiments in \cite{CanedoETAL}, the $x-z$ cross section shows minimal overall bubble rise speed in the (very) viscous fluid in comparison to the rate that the bubble deforms in the $y-z$ cross section.  This is expected since the experiments designed by \cite{CanedoETAL} closely followed the following basic assumptions: (i) ``steady creeping flow with negligible inertial effects,'' (ii) ``Incompressible Newtonian fluids,'' (iii) ``No buoyancy effects,'' (iv) ``No wall effects.''  In this work, we determine, for the first time, the critical Reynolds number ($Re \gg 1$) that leads to bubble breakup.  Additionally, our computational studies reveal characteristics that distinguish a drop's deformation and breakup processes versus those of a bubble.

@@

\subsection{Numerical method and governing equations}
%  -----------------------------------------------------------------------------
Numerical results were obtained using the interface capturing Coupled Level Set and Volume of Fluid (CLSVOF) method (\citet{SusPuc00,SusSmiHusOhtZhi07}), which is based on a fixed grid finite volume algorithm.  The CLSVOF method is a robust  numerical  technique that  combines  some  of  the  advantages  of the Volume of Fluid (VOF) method (\citet{HirNic81}) and the Level Set (LS) (\citet{SusSmeOsh94}) method while overcoming their weaknesses.  In the VOF method,  the Volume Fraction function, $F$, is used to represent the interface. The values of $F$ correspond to the volume fraction of liquid in a given computational cell.  In other words, $F = 0$ when a computational cell contains only gas and $F = 1$ when a computational cell contains only liquid.  If $0 < F < 1$, then a computational cell contains the gas-liquid interface. The VOF method has a great advantage over the LS method in that accurate algorithms for advecting $F$ can be applied so that mass/volume is conserved up to machine precision while still maintaining a sharp representation of the interface.  On the other hand, the disadvantage of the VOF method in comparison to the LS method is that tangled and difficult reconstruction procedures are required for determining the slope of the piecewise linear VOF reconstructed interface.  In the LS method, the signed distance function $\phi$ (LS function) is used to track the interface. The interface is implicitly represented by the set of points in which $\phi = 0$.  Liquid and gas regions are defined as $\phi > 0$ in the liquid and $\phi < 0$ in the gas, respectively.  One of the advantages of the LS method is that one can track and represent smoothly the interface, but the LS method has the disadvantage that mass/volume is not explicitly conserved.  In the CLSVOF method, the coupling between the LS function and the VOF function occurs when computing the normal of the reconstructed interface in the VOF calculation process and also when assigning the LS function with the exact signed normal distance to the reconstructed interface in the LS calculation process. That is to say, the piecewise linear approximation (the volume-of-fluid reconstruction step) for the VOF method is determined using the unit normal vector ($\bm n$) estimated from information of the LS function. By taking advantage of both methods, the evolution of the liquid-gas interface location can be computationally captured in such a way so that volume/mass is preserved to machine precision and at the same time, the interface normals and the surface tension force (which is proportional to the interface curvature) can be straightforwardly derived from the smooth level set function.
\par 
In our studies, the two-phase fluid flow is comprised of air and a viscous Newtonian liquid.  The Heaviside function, $\Hea(\phi)$, which is defined as
\begin{equation}\label{heavyeqn}
  \Hea(\phi) = \begin{cases}
               1 & \phi \geq 0 \\
               0 & \phi <0 
               \end{cases}
\end{equation}
will be used below to distinguish each of the two fluids.  A single set of
three-dimensional equations governs the motion of both fluids, which are taken
to be incompressible, and consists of the continuity equation and the
Navier-Stokes equations with surface tension forces:
\begin{align}
  \nabla\cdot\bmu &=0  \label{divu} \\
  \frac{\partial\bmu}{\partial t}+(\bmu\cdot\nabla)\bmu &=
  \frac{1}{\rho}\nabla\cdot(-p\bmI+2\mu\bmD)+\bmg-
  \frac{\sigma\kappa}{\rho}\nabla \Hea  
  \label{nseqn}
\end{align}

@@

@@

The numerical validation studies performed in this section and the
following section demonstrate that our numerical method can reliably determine
the transition regions at which shear-induced bubble or drop deformation leads
to breakup.  We remark that in the next section, we demonstrate that we can
expect an error of $3\%$ for predicting the transition to break-up.  The
analysis in this section and the following also demonstrate that the error is
reduced by a factor of 2 each time the grid is refined by a factor of 2. Also,
we refer the reader to \citet{ohta2011robust} in which we apply our CLSVOF
method for bubble formation problems.

@@

%  -----------------------------------------------------------------------------
\section{Conclusions}
%  -----------------------------------------------------------------------------
The bubble deformation and breakup process in simple linear shear flow liquid
was explored numerically using the CLSVOF computational method.  In this study,
the critical Reynolds number $Re_{\rm c}$, at which bubble breakup first occurs,
was determined for several flow conditions, and the differences between bubble
deformation and breakup were compared with the well-known analogous process of
drop deformation and breakup.

Numerical results revealed significant differences between bubble deformation
and breakup and the corresponding drop dynamics.  For case of bubble, it was
discovered that much stronger shear flows are necessary to induce interface
breakup compared with a drop immersed in a similar flow field.  That is, a much
larger Reynolds number flow is required in order to induce bubble breakup.  The
behavior of bubble breakup was very similar through the $Ca$ number range
considered in our computations: the bubble underwent a similar breakup
mechanism in which rupture occurred at a thread-bridge part that followed a
doglegged shape formation stage.  In bubble deformation without breakup, near
$Re_{\rm c}$, the bubble did not maintain a stable deformed shape, in contrast to
drop deformation near the critical Reynolds number.  The bubble exhibited
pronounced underdamped behavior: the bubble oscillated between elongating and
shrinking motions for non-rupturing flow conditions.
\textcolor{red}
{
At the same time, bubble deformation under smaller $Re$ conditions ($< Re_{\rm c}$)
becomes in a stable state.
}
We attribute the large differences in morphology for the bubble undergoing
breakup, compared with the drop, to the density and viscosity ratio.  The
density and viscosity ratio remarkably impacts on bubble/drop deformation and
breakup.  The bubble deformation and breakup is subject to a synergistic
coupling of the density and viscosity ratio, and whose effect will be examined
separately in future work.

>>>>

\begin{abstract}
\textcolor{red}
{
Numerical simulations are presented for the deformation and breakup of a bubble in liquids undergoing simple linear shear flow in a variety of regimes including the low Capillary number, high Reynolds number regime.  Numerical results are obtained using a projection method for incompressible two-phase flow. The method represents interfaces using the sharp interface coupled level set and Volume-Of-Fluid (CLSVOF) method.   To verify the CLSVOF numerical algorithm and provide a basis for comparison, computational results are also presented that compare with previous, more popular, drop deformation and rupture simulations.  The numerical results reveal that the shear-induced bubble deformation and breakup process differs significantly from that of the analogous drop system under similar flow conditions.  The first distinguishing factor for bubble breakup is that the magnitude of shear flow necessary for causing rupture is much larger than that in the drop system case.  In other words, a larger Reynolds number is needed to induce bubble breakup.  The second distinct feature of the bubble system versus the drop system is that the bubble does not maintain a stable deformed shape as the system parameters approach the critical ``breakup'' Reynolds number.  It is asserted that the differences in morphology for a bubble undergoing breakup versus a drop in the same process can be attributed to the density and viscosity ratio of the corresponding two-phase flow systems.  The critical conditions for bubble breakup with respect to the Reynolds and capillary numbers are determined for several cases. }
\end{abstract}

@@

Bubble dynamics in shear flow, including breakup, is critically important for various \textcolor{blue} {gas-liquid} scientific and engineering processes.  We refer the reader to the following experimental studies relating to bubble deformation in foaming processes, microfluidic devices, microbubbles in the blood circulation system, materials processing, pharmaceuticals, minerals engineering, bubble column reactor, underwater projectiles, polyethylene devolatilization, microfluidics, food aeration, passive cyclonic separators, and cosmetics\cite{ChuFinBouAtaHamPug19,MulTobDreFisWin08,BenRodFauPinFerPerGarMirLim18,DreSai15,EFTEKHARI2021837,WANG2023108105,doi:10.1021/acs.langmuir.1c01814,yoshikawa2010bubble,CanedoETAL,GAO2022103212,WONG2012417,PolyethyleneDevo,FRENSE2024120579,lohse2018bubble,schluter2021small,SANOGO2023112478,hoyt2013performance,sines2020study}.  In particular, it is the study of bubble deformation as it pertains to high-performance plastics applications that motivate this work.

@@

In previous experimental or computational studies on the motion of bubble deformation in a simple shear flow \cite{CanedoETAL,RusMan02, MulTobDreFisWin08}, only findings for bubble deformation under very low $Re$ number conditions ($Re \ll 1$) have been reported.  This is understandable since a low Reynolds number matrix fluid mitigates the effect of gravity on distorting the comparison with the drop case.  In the results reported in \cite{CanedoETAL}, the $x-z$ cross section of the deforming bubble was close to circular in their experiments.  Also in the experiments in \cite{CanedoETAL}, the $x-z$ cross section shows minimal overall bubble rise speed in the (very) viscous fluid in comparison to the rate that the bubble deforms in the $y-z$ cross section.  This is expected since the experiments designed by \cite{CanedoETAL} closely followed the following basic assumptions: (i) ``steady creeping flow with negligible inertial effects,'' (ii) ``Incompressible Newtonian fluids,'' (iii) ``No buoyancy effects,'' (iv) ``No wall effects.''  In this work, we determine, for the first time, the critical Reynolds number ($Re \gg 1$) that leads to bubble breakup.  Additionally, our computational studies reveal characteristics that distinguish a drop's deformation and breakup processes versus those of a bubble.

@@

\subsection{Numerical method and governing equations}
%  -----------------------------------------------------------------------------
Numerical results were obtained using the interface capturing Coupled Level Set and Volume of Fluid (CLSVOF) method (\citet{SusPuc00,SusSmiHusOhtZhi07}), which is based on a fixed grid finite volume algorithm.  The CLSVOF method is a robust  numerical  technique that  combines  some  of  the  advantages  of the Volume of Fluid (VOF) method (\citet{HirNic81}) and the Level Set (LS) (\citet{SusSmeOsh94}) method while overcoming their weaknesses.  In the VOF method,  the Volume Fraction function, $F$, is used to represent the interface. The values of $F$ correspond to the volume fraction of liquid in a given computational cell.  In other words, $F = 0$ when a computational cell contains only gas and $F = 1$ when a computational cell contains only liquid.  If $0 < F < 1$, then a computational cell contains the gas-liquid interface. The VOF method has an advantage over the LS method in that accurate algorithms for advecting $F$ can be applied so that mass/volume is conserved up to machine precision while still maintaining a sharp representation of the interface.  On the other hand, the disadvantage of the VOF method in comparison to the LS method is that tangled and difficult reconstruction procedures are required for determining the slope of the piecewise linear VOF reconstructed interface.  In the LS method, the signed distance function $\phi$ (LS function) is used to track the interface. The interface is implicitly represented by the set of points in which $\phi = 0$.  Liquid and gas regions are defined as $\phi > 0$ in the liquid and $\phi < 0$ in the gas, respectively.  One of the advantages of the LS method is that one can track and represent smoothly the interface, but the LS method has the disadvantage that mass/volume is not explicitly conserved.  In the CLSVOF method, the coupling between the LS function and the VOF function occurs when computing the normal of the reconstructed interface in the VOF calculation process and also when assigning the LS function with the exact signed normal distance to the reconstructed interface in the LS calculation process. That is to say, the piecewise linear approximation (the volume-of-fluid reconstruction step) for the VOF method is determined using the unit normal vector ($\bm n$) estimated from information of the LS function. By taking advantage of both methods, the evolution of the liquid-gas interface location can be computationally captured in such a way so that volume/mass is preserved to machine precision and at the same time, the interface normals and the surface tension force (which is proportional to the interface curvature) can be straightforwardly derived from the smooth level set function.
\par 
In our studies, the two-phase fluid flow is comprised of air and a viscous Newtonian liquid.  The Heaviside function, $\Hea(\phi)$, which is defined as
\begin{equation}\label{heavyeqn}
  \Hea(\phi) = \begin{cases}
               1 & \phi \geq 0 \\
               0 & \phi <0 
               \end{cases}
\end{equation}
will be used below to distinguish each of the two fluids.  A single set of
three-dimensional equations governs the motion of both fluids, which are taken
to be incompressible, and consists of the continuity equation and the
Navier-Stokes equations with surface tension forces:
\begin{align}
  \nabla\cdot\bmu &=0  \label{divu} \\
  \frac{\partial\bmu}{\partial t}+(\bmu\cdot\nabla)\bmu &=
  \frac{1}{\rho}\nabla\cdot(-p\bmI+2\mu\bmD)+\bmg-
  \frac{\sigma\kappa}{\rho}\nabla \Hea  
  \label{nseqn}
\end{align}

@@

\textcolor{red}
{
Remark 1: Due to time step stability constraints, the variable density pressure projection process, and computed bubble shapes with high aspect ratio, we find that our simulations can take over six months.  We have experimented with (a) decreasing the ``error buffer'' parameter (radius of cells to be tagged when a given cell is tagged for adaptivity) and (b) relaxing the condition that the bubble-liquid interface be wholly contained on the finest adaptive level.  Unfortunately we have found that these steps lead to poorer accuracy.  This ``diminishing returns'' phenomena is expected for low Mach number flows in which the incompressible flow equations are characterized by non-local behavior.  We refer the reader to the following research\cite{li2016high} in which it has been found through a systematic study that using an AMR grid can be less accurate than a case with a uniform fine grid (luckily that is not the case here).  To summarize: we have found that each further refinement of the grid will multiply the simulation time by a factor of about eight (factor of 4 due to spatial refinement, factor of 2 due to temporal refinement).
}

\textcolor{red}
{
Remark 2: We believe that the inclusion of a special sub-scale model right at the point of bubble break-up is unnecessary because the driving shear force is temporally uniformly applied instead of impulsively applied.  We are aware of research for predicting whether droplets merge or bounce\cite{Lewin-Jones_Lockerby_Sprittles_2024} but this research is not applicable in our case.  We are not aware of previous studies on the shear flow driven breakup of bubble or drops that implement special subscale models.
}

@@

\textcolor{red}{
The numerical validation studies performed in this section and the
following section demonstrate that our numerical method can reliably determine
the transition regions at which shear-induced bubble or drop deformation leads
to breakup.  We remark that in the next section, we demonstrate that we can
expect an error of $3\%$ for predicting the transition to break-up.  The
analysis in this section and the following also demonstrate that the error is
reduced by a factor of 2 each time the grid is refined by a factor of 2. We reiterate that we have found at least a factor of 2 error reduction for each grid refinement in a number of multiphase flow problems involving complex interface deformation and breakup; see \citet{OhtSus12, OhtAkaYosSus14, OhtFurYosSus19,ohta2010sensitivity,stewart2008improved,SusSmiHusOhtZhi07,arienti2013coupled,ohta2011robust}.
}

@@

%  -----------------------------------------------------------------------------
\section{Conclusions}
%  -----------------------------------------------------------------------------
\textcolor{red}{
The bubble deformation and breakup process in liquid due to a driving simple linear shear flow was explored numerically using the CLSVOF computational method.  In this study, the critical Reynolds number $Re_{\rm c}$, at which bubble breakup first occurs, was determined for several flow conditions, and the differences between the morphology of bubble deformation and breakup were compared with the analogous morphology of drop deformation and breakup.
}

\textcolor{red}{
The numerical results revealed significant differences between bubble deformation and breakup and the corresponding drop dynamics.  For the case of a bubble, it was discovered that much stronger shear flows are necessary in order to induce interface breakup compared with a drop immersed in a similar flow field.  That is, a much larger Reynolds number flow is required in order to induce bubble breakup.  The steps leading to bubble breakup were similar throughout the $Ca$ number range considered in our computations: the bubble underwent a similar breakup mechanism in which rupture occurred at a thread-bridge part that followed a doglegged shape formation stage.  For bubble deformation without breakup, near $Re_{\rm c}$, the bubble did not maintain a stable deformed shape, in contrast to drop deformation near the critical Reynolds number.  The bubble exhibited pronounced underdamped behavior: the bubble oscillated between elongating and shrinking motions for non-rupturing flow conditions.  At the same time, bubble deformation under smaller $Re$ conditions ($< Re_{\rm c}$) led to a stable state.  We attribute the large differences in morphology for the bubble undergoing breakup, compared with the drop, due to the density and viscosity ratio.  
}

\textcolor{red}{
There are a number of directions for future work.  In the present work, we varied the Capillary number and Reynolds' number in studying bubble deformation and breakup morphology in the presence of a simple shear flow.  For the future, one can also vary (i) the viscosity ratio, (ii) the effect of the initial shape of the bubble\cite{ohta2005computational}, (iii) the magnitude and direction of the gravitational force, (iv) the non-Newtonian properties of the ``matrix'' liquid, (v) bubble deformation and breakup due to a shear flow in a T junction\cite{FRENSE2024120579}, (vi) bubble deformation and breakup confined shear flow (see \cite{gupta2014deformation} for the drop case), and (vii) bubble deformation and breakup due to non-uniform shear flow (see \cite{chin1980studies} for the drop case).
}
